---
title: "EHH Haplotype-Based Selection Analysis — Chromosome 3"
subtitle: "P. aegeria backcross GWAS: NN vs DD populations"
author: "GWAS Pipeline"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    code_folding: show
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
```

# Background

## Extended Haplotype Homozygosity (EHH)

EHH-based statistics detect **signatures of recent positive selection** by examining the
decay of haplotype homozygosity around a focal variant. Under neutrality, haplotypes
break down quickly due to recombination. Under positive selection, a beneficial allele
sweeps to high frequency carrying a long, intact haplotype — creating an unusually
extended region of homozygosity.

### Key statistics

| Statistic | What it detects | Comparison |
|-----------|----------------|------------|
| **iHS** (integrated haplotype score) | Ongoing/incomplete selective sweeps within a single population | Compares EHH of ancestral vs derived allele at same locus |
| **XP-EHH** (cross-population EHH) | Completed or near-complete sweeps that differ between populations | Compares haplotype lengths between two populations |
| **Rsb** | Similar to XP-EHH but uses iES (integrated EHHS) | Between-population comparison |

### Important caveats

- Data **must be phased** for EHH statistics to be meaningful. We use **Beagle 5.5** to
  statistically phase our FreeBayes VCF before running `rehh`.
- We use `polarize_vcf = FALSE` since we lack an outgroup to define ancestral/derived
  alleles. The package uses major/minor allele frequency instead.
- EHH methods assume a recombination map is available. Without one, physical positions
  (bp) are used as a proxy. Beagle assumes a uniform rate of 1 cM/Mb when no genetic
  map is provided — this is approximate but standard for non-model organisms.

**Note on phasing tool choice:** SHAPEIT5 was our first choice but is only available
for Linux (x86_64), not macOS ARM64. Beagle 5.5 is Java-based (cross-platform),
widely used for non-model organisms, and produces comparable phasing accuracy —
especially at moderate sample sizes (~84 diploid individuals) without a reference panel.

## Why phasing matters

EHH statistics measure the extent of **haplotype** homozygosity around a focal SNP.
Without phasing, we don't know which alleles are on the same chromosome. Unphased data
forces `rehh` to treat each diploid genotype as two independent haplotypes with random
phase assignment, which:

- Underestimates the true extent of haplotype homozygosity
- Reduces power to detect selective sweeps
- Can create false signals from phase-switching artefacts

Beagle uses LD patterns across the 84 diploid samples (168 haplotypes) to
statistically infer phase. With no genetic map available for *P. aegeria*, it
assumes a uniform recombination rate of 1 cM/Mb.

# Setup

## Install and load packages

```{r install-packages, eval=FALSE}
# Run once if not installed
install.packages("rehh")
install.packages("tidyverse")
install.packages("data.table")
```

```{r load-packages}
library(rehh)
library(tidyverse)
library(data.table)
```

## Define paths

```{r paths}
# Project paths
project_dir <- "/Users/stockholmbutterflylab/sbl_claudecode/Annas_GWAS"
vcf_full    <- file.path(project_dir, "results/BayPass.gwas_maf0.1geno0.2/vcf/freebayes_filtered.vcf.gz")
samples_tsv <- file.path(project_dir, "samples.tsv")

# Tool paths (from Snakemake conda environments)
conda_bin   <- file.path(project_dir, ".snakemake/conda/aedce8cd90aec491ec84dcf2bee0648f_/bin")
bcftools    <- file.path(conda_bin, "bcftools")

# Output directory for intermediate files
out_dir <- file.path(project_dir, "results/EHH_analysis")
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

# Chromosome of interest
chr_target <- "3"
```

## Phasing tool: Beagle 5.5

Beagle 5.5 is installed via micromamba in a dedicated environment.
It requires Java (OpenJDK), which is included in the environment.

```{r install-beagle, eval=FALSE}
# Run once to install Beagle (already done if following this pipeline)
system("MAMBA_ROOT_PREFIX=~/micromamba /tmp/bin/micromamba create -n phasing -c bioconda -c conda-forge beagle openjdk -y")
```

```{r find-beagle}
# micromamba runner for the phasing environment
micromamba <- "/tmp/bin/micromamba"
mamba_prefix <- file.path(Sys.getenv("HOME"), "micromamba")

# Helper function to run commands in the phasing environment
run_beagle_env <- function(cmd) {
  full_cmd <- sprintf("MAMBA_ROOT_PREFIX=%s %s run -n phasing bash -c '%s'",
                       mamba_prefix, micromamba, cmd)
  system(full_cmd, intern = TRUE)
}

# Verify Beagle is accessible
beagle_test <- run_beagle_env("beagle 2>&1 | head -1")
cat("Beagle version:", beagle_test, "\n")
```

# Data Preparation

## Read sample metadata

```{r sample-metadata}
samples <- read_tsv(samples_tsv, show_col_types = FALSE)

# Get the actual samples present in the VCF
vcf_samples <- system(sprintf("%s query -l %s 2>/dev/null", bcftools, vcf_full), intern = TRUE)
cat("Samples in VCF:", length(vcf_samples), "\n")
cat("Samples in metadata:", nrow(samples), "\n")

# Filter metadata to only samples present in the VCF
samples <- samples %>% filter(sample %in% vcf_samples)
cat("Samples matched:", nrow(samples), "\n\n")

# Summarise groups
cat("Sample groups:\n")
table(samples$group)

# Create sample lists for each population
nn_samples <- samples %>% filter(group == "NN") %>% pull(sample)
dd_samples <- samples %>% filter(group == "DD") %>% pull(sample)

cat("\nNN samples:", length(nn_samples), "\n")
cat("DD samples:", length(dd_samples), "\n")
```

## Step 1: Extract chromosome 3 (biallelic SNPs)

We use `bcftools` to extract biallelic SNPs on chromosome 3 and prepare
the VCF for phasing.

```{r extract-chr3, results='hide'}
# Write sample lists to files
nn_list_file <- file.path(out_dir, "nn_samples.txt")
dd_list_file <- file.path(out_dir, "dd_samples.txt")
writeLines(nn_samples, nn_list_file)
writeLines(dd_samples, dd_list_file)

# Step 1a: Extract biallelic SNPs on chr3 (all samples)
vcf_chr3_raw <- file.path(out_dir, "chr3_biallelic_snps.vcf.gz")
cmd_extract <- sprintf(
  "%s view -r %s -m2 -M2 -v snps --min-af 0.05:minor %s -Oz -o %s && %s index %s",
  bcftools, chr_target, vcf_full, vcf_chr3_raw,
  bcftools, vcf_chr3_raw
)
cat("Extracting chr3 biallelic SNPs...\n")
system(cmd_extract)

# Step 1b: Fix ploidy — FreeBayes can produce haploid genotype calls at some sites
# (e.g., "0" instead of "0/0"). Beagle requires all genotypes to be diploid.
vcf_chr3_diploid <- file.path(out_dir, "chr3_biallelic_snps_diploid.vcf.gz")
cmd_fixploidy <- sprintf(
  "%s +fixploidy %s -Oz -o %s -- -f 2 && %s index %s",
  bcftools, vcf_chr3_raw, vcf_chr3_diploid,
  bcftools, vcf_chr3_diploid
)
cat("Fixing ploidy (haploid → diploid)...\n")
system(cmd_fixploidy)

cat("VCF extraction and ploidy fix complete.\n")
```

```{r verify-extraction}
cmd <- sprintf("%s stats %s 2>/dev/null | grep '^SN'", bcftools, vcf_chr3_diploid)
result <- system(cmd, intern = TRUE)
cat("Chromosome 3 stats (pre-phasing):\n")
cat(paste(result, collapse = "\n"), "\n")
```

## Step 2: Phase with Beagle 5.5

We run Beagle on all 84 samples together. Phasing jointly gives the best
statistical power — we split into populations afterwards.

**Key parameters:**

- No genetic map available → uniform 1 cM/Mb assumed (default)
- No reference panel → phasing from LD within our 84 samples only
- `impute=false` → we only want phasing, not imputation of ungenotyped sites
- `chrom=3` → restrict to chromosome 3
- `nthreads=8` for parallel processing

```{r run-beagle}
vcf_chr3_phased <- file.path(out_dir, "chr3_phased.vcf.gz")
beagle_prefix   <- file.path(out_dir, "chr3_phased")

# Only run if phased file doesn't already exist
if (!file.exists(vcf_chr3_phased)) {

  # Use the ploidy-fixed VCF as input; -Xmx8g for sufficient heap space
  beagle_cmd <- sprintf(
    "beagle -Xmx8g gt=%s out=%s chrom=%s impute=false nthreads=8 2>&1",
    vcf_chr3_diploid, beagle_prefix, chr_target
  )

  cat("Running Beagle 5.5 phasing...\n")
  cat("Command:", beagle_cmd, "\n\n")

  phase_result <- run_beagle_env(beagle_cmd)
  cat(paste(tail(phase_result, 20), collapse = "\n"), "\n")

  # Index the phased output
  system(sprintf("%s index -t %s", bcftools, vcf_chr3_phased))
  cat("\nPhasing complete.\n")
} else {
  cat("Phased file already exists, skipping Beagle.\n")
}
```

## Step 3: Verify phasing

```{r verify-phasing}
# Check that genotypes are now phased (pipe-separated: 0|1 instead of 0/1)
cmd_check <- sprintf(
  "%s query -f '%%POS\\t[%%GT\\t]\\n' %s 2>/dev/null | head -5",
  bcftools, vcf_chr3_phased
)
phased_check <- system(cmd_check, intern = TRUE)
cat("First 5 phased genotypes (should show 0|1, 1|0, etc.):\n")
for (line in phased_check) {
  # Show first few fields only
  fields <- strsplit(line, "\t")[[1]]
  cat("  pos", fields[1], ":", paste(fields[2:min(6, length(fields))], collapse = " "), "...\n")
}

# Count phased vs unphased
cmd_count <- sprintf(
  "%s query -f '[%%GT\\n]' %s 2>/dev/null | head -10000 | grep -c '|'",
  bcftools, vcf_chr3_phased
)
n_phased <- as.integer(system(cmd_count, intern = TRUE))
cat("\nPhased genotypes in first 10000:", n_phased, "/ 10000\n")
```

## Step 4: Split phased VCF by population

```{r split-populations, results='hide'}
vcf_chr3_nn <- file.path(out_dir, "chr3_phased_NN.vcf.gz")
vcf_chr3_dd <- file.path(out_dir, "chr3_phased_DD.vcf.gz")

# NN population
cmd_nn <- sprintf(
  "%s view -S %s %s -Oz -o %s && %s index %s",
  bcftools, nn_list_file, vcf_chr3_phased, vcf_chr3_nn,
  bcftools, vcf_chr3_nn
)
system(cmd_nn)

# DD population
cmd_dd <- sprintf(
  "%s view -S %s %s -Oz -o %s && %s index %s",
  bcftools, dd_list_file, vcf_chr3_phased, vcf_chr3_dd,
  bcftools, vcf_chr3_dd
)
system(cmd_dd)

cat("Population-split phased VCFs created.\n")
```

## Verify final VCFs

```{r verify-vcfs}
for (vcf_path in c(vcf_chr3_nn, vcf_chr3_dd)) {
  cmd <- sprintf("%s stats %s 2>/dev/null | grep '^SN.*number of SNPs'", bcftools, vcf_path)
  result <- system(cmd, intern = TRUE)
  cat(basename(vcf_path), ":", result, "\n")
}
```

# Load phased data into rehh

```{r load-rehh}
# Load ALL samples combined (phased VCF)
all_hh <- data2haplohh(
  hap_file = vcf_chr3_phased,
  polarize_vcf = FALSE,
  vcf_reader = "data.table",
  chr.name = chr_target,
  verbose = FALSE
)

# Load NN population (phased VCF)
nn_hh <- data2haplohh(
  hap_file = vcf_chr3_nn,
  polarize_vcf = FALSE,
  vcf_reader = "data.table",
  chr.name = chr_target,
  verbose = FALSE
)

# Load DD population (phased VCF)
dd_hh <- data2haplohh(
  hap_file = vcf_chr3_dd,
  polarize_vcf = FALSE,
  vcf_reader = "data.table",
  chr.name = chr_target,
  verbose = FALSE
)

cat("ALL haplohh object:", nrow(all_hh@haplo), "haplotypes x", ncol(all_hh@haplo), "markers\n")
cat("NN haplohh object:", nrow(nn_hh@haplo), "haplotypes x", ncol(nn_hh@haplo), "markers\n")
cat("DD haplohh object:", nrow(dd_hh@haplo), "haplotypes x", ncol(dd_hh@haplo), "markers\n")
```

## Filter for minor allele frequency

```{r maf-filter}
all_hh_f <- subset(all_hh, min_maf = 0.05)
nn_hh_f <- subset(nn_hh, min_maf = 0.05)
dd_hh_f <- subset(dd_hh, min_maf = 0.05)

cat("After MAF filter:\n")
cat("ALL:", ncol(all_hh_f@haplo), "markers\n")
cat("NN:", ncol(nn_hh_f@haplo), "markers\n")
cat("DD:", ncol(dd_hh_f@haplo), "markers\n")
```

# iHS Selection Scan

## Compute EHH scans

`scan_hh()` computes integrated EHH (iHH) for each SNP — the area under the EHH
decay curve for each allele. This is the raw material for iHS, XP-EHH and Rsb.

```{r scan-hh}
all_scan <- scan_hh(all_hh_f, polarized = FALSE)
nn_scan  <- scan_hh(nn_hh_f, polarized = FALSE)
dd_scan  <- scan_hh(dd_hh_f, polarized = FALSE)

cat("ALL scan:", nrow(all_scan), "variants scanned\n")
cat("NN scan: ", nrow(nn_scan), "variants scanned\n")
cat("DD scan: ", nrow(dd_scan), "variants scanned\n")
```

## Compute iHS

`ihh2ihs()` converts the raw iHH values to the standardised iHS statistic.
Since our data is unpolarised, we use `freqbin = 1` (single frequency bin)
to avoid frequency-dependent standardisation artefacts.

```{r compute-ihs}
all_ihs <- ihh2ihs(all_scan, freqbin = 1)
nn_ihs  <- ihh2ihs(nn_scan, freqbin = 1)
dd_ihs  <- ihh2ihs(dd_scan, freqbin = 1)
```

## Plot iHS — All samples combined {.tabset}

The combined iHS treats all 84 individuals (NN + DD) as a single population.
This maximises sample size and haplotype diversity, giving the best power to
detect sweeps shared across the backcross populations.

### iHS values & p-values

```{r plot-ihs-all, fig.height=8}
par(mfrow = c(2, 1))

plot(all_ihs$ihs$POSITION / 1e6, all_ihs$ihs$IHS,
     pch = 16, cex = 0.3, col = alpha("grey30", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "iHS",
     main = "iHS — All samples combined (Chromosome 3)")
abline(h = c(-2, 2), col = "red", lty = 2)

plot(all_ihs$ihs$POSITION / 1e6, all_ihs$ihs$LOGPVALUE,
     pch = 16, cex = 0.3, col = alpha("grey30", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "-log10(p-value)",
     main = "iHS -log10(p) — All samples combined")
abline(h = 4, col = "red", lty = 2)
abline(h = 6, col = "darkred", lty = 2)

par(mfrow = c(1, 1))
```

### Top 20 combined iHS hits

```{r top-ihs-all}
all_ihs$ihs %>%
  as_tibble() %>%
  arrange(desc(LOGPVALUE)) %>%
  head(20) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3)) %>%
  select(CHR, POSITION, POSITION_Mb, IHS, LOGPVALUE)
```

## Plot iHS — Per population {.tabset}

### NN population

```{r plot-ihs-nn, fig.height=8}
par(mfrow = c(2, 1))

plot(nn_ihs$ihs$POSITION / 1e6, nn_ihs$ihs$IHS,
     pch = 16, cex = 0.3, col = alpha("steelblue", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "iHS",
     main = "iHS — NN population (Chromosome 3)")
abline(h = c(-2, 2), col = "red", lty = 2)

plot(nn_ihs$ihs$POSITION / 1e6, nn_ihs$ihs$LOGPVALUE,
     pch = 16, cex = 0.3, col = alpha("steelblue", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "-log10(p-value)",
     main = "iHS -log10(p) — NN population")
abline(h = 4, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

### DD population

```{r plot-ihs-dd, fig.height=8}
par(mfrow = c(2, 1))

plot(dd_ihs$ihs$POSITION / 1e6, dd_ihs$ihs$IHS,
     pch = 16, cex = 0.3, col = alpha("darkorange", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "iHS",
     main = "iHS — DD population (Chromosome 3)")
abline(h = c(-2, 2), col = "red", lty = 2)

plot(dd_ihs$ihs$POSITION / 1e6, dd_ihs$ihs$LOGPVALUE,
     pch = 16, cex = 0.3, col = alpha("darkorange", 0.5),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "-log10(p-value)",
     main = "iHS -log10(p) — DD population")
abline(h = 4, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

## Top iHS hits — per population

```{r top-ihs}
cat("=== Top 20 iHS hits — NN population ===\n")
nn_ihs$ihs %>%
  as_tibble() %>%
  arrange(desc(LOGPVALUE)) %>%
  head(20) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3)) %>%
  select(CHR, POSITION, POSITION_Mb, IHS, LOGPVALUE)

cat("\n=== Top 20 iHS hits — DD population ===\n")
dd_ihs$ihs %>%
  as_tibble() %>%
  arrange(desc(LOGPVALUE)) %>%
  head(20) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3)) %>%
  select(CHR, POSITION, POSITION_Mb, IHS, LOGPVALUE)
```

# Cross-Population Selection Scan (XP-EHH)

XP-EHH compares haplotype lengths between two populations. Positive values indicate
selection in the first population (DD); negative values indicate selection in the
second population (NN).

## Compute XP-EHH

```{r compute-xpehh}
xpehh <- ies2xpehh(
  dd_scan, nn_scan,
  popname1 = "DD",
  popname2 = "NN",
  include_freq = TRUE
)

cat("XP-EHH computed for", nrow(xpehh), "variants\n")
```

## Plot XP-EHH

```{r plot-xpehh, fig.height=10}
par(mfrow = c(2, 1))

# XP-EHH values
plot(xpehh$POSITION / 1e6, xpehh$XPEHH_DD_NN,
     pch = 16, cex = 0.3,
     col = ifelse(xpehh$XPEHH_DD_NN > 0, alpha("darkorange", 0.4), alpha("steelblue", 0.4)),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "XP-EHH (DD vs NN)",
     main = "XP-EHH — DD vs NN (Chromosome 3)")
abline(h = 0, col = "grey40", lty = 1)
abline(h = c(-2, 2), col = "red", lty = 2)
legend("topright",
       legend = c("Selection in DD (+)", "Selection in NN (-)"),
       col = c("darkorange", "steelblue"), pch = 16, cex = 0.8)

# Log p-values
plot(xpehh$POSITION / 1e6, xpehh$LOGPVALUE,
     pch = 16, cex = 0.3, col = alpha("purple", 0.4),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "-log10(p-value)",
     main = "XP-EHH -log10(p) — DD vs NN")
abline(h = 4, col = "red", lty = 2)
abline(h = 6, col = "darkred", lty = 2)

par(mfrow = c(1, 1))
```

## Top XP-EHH hits

```{r top-xpehh}
xpehh_tbl <- xpehh %>%
  as_tibble() %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3))

cat("=== Top 20 XP-EHH hits (selection in DD, positive values) ===\n")
xpehh_tbl %>%
  arrange(desc(XPEHH_DD_NN)) %>%
  head(20) %>%
  select(CHR, POSITION, POSITION_Mb, XPEHH_DD_NN, LOGPVALUE, starts_with("FREQ_A1"))

cat("\n=== Top 20 XP-EHH hits (selection in NN, negative values) ===\n")
xpehh_tbl %>%
  arrange(XPEHH_DD_NN) %>%
  head(20) %>%
  select(CHR, POSITION, POSITION_Mb, XPEHH_DD_NN, LOGPVALUE, starts_with("FREQ_A1"))

cat("\n=== Top 20 by p-value (either direction) ===\n")
xpehh_tbl %>%
  arrange(desc(LOGPVALUE)) %>%
  head(20) %>%
  select(CHR, POSITION, POSITION_Mb, XPEHH_DD_NN, LOGPVALUE)
```

# Rsb (between-population iES comparison)

Rsb provides an alternative cross-population test using standardised integrated EHHS.

```{r compute-rsb}
rsb <- ines2rsb(
  dd_scan, nn_scan,
  popname1 = "DD",
  popname2 = "NN"
)

cat("Rsb computed for", nrow(rsb), "variants\n")
```

```{r plot-rsb, fig.height=8}
par(mfrow = c(2, 1))

plot(rsb$POSITION / 1e6, rsb$RSB_DD_NN,
     pch = 16, cex = 0.3,
     col = ifelse(rsb$RSB_DD_NN > 0, alpha("darkorange", 0.4), alpha("steelblue", 0.4)),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "Rsb (DD vs NN)",
     main = "Rsb — DD vs NN (Chromosome 3)")
abline(h = 0, col = "grey40")
abline(h = c(-2, 2), col = "red", lty = 2)

plot(rsb$POSITION / 1e6, rsb$LOGPVALUE,
     pch = 16, cex = 0.3, col = alpha("purple", 0.4),
     xlab = "Position on Chromosome 3 (Mb)",
     ylab = "-log10(p-value)",
     main = "Rsb -log10(p) — DD vs NN")
abline(h = 4, col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Haplotype Structure at Top Hits

## Identify top XP-EHH hit

```{r find-top-hit}
top_hit <- xpehh_tbl %>% arrange(desc(LOGPVALUE)) %>% slice(1)

cat("Top XP-EHH hit:\n")
cat("  Position:", top_hit$POSITION, "(", round(top_hit$POSITION / 1e6, 3), "Mb)\n")
cat("  XP-EHH:", round(top_hit$XPEHH_DD_NN, 3), "\n")
cat("  -log10(p):", round(top_hit$LOGPVALUE, 3), "\n")

focal_pos <- top_hit$POSITION
```

## Bifurcation diagrams

Bifurcation diagrams show how haplotypes branch away from a focal SNP.
Under selection, one haplotype extends much further than the other.

```{r bifurcation, fig.height=8}
# Find marker index in each haplohh object
marker_nn <- which(nn_hh_f@positions == focal_pos)
marker_dd <- which(dd_hh_f@positions == focal_pos)

if (length(marker_nn) > 0 & length(marker_dd) > 0) {

  # Compute furcation trees
  furc_nn <- calc_furcation(nn_hh_f, mrk = marker_nn)
  furc_dd <- calc_furcation(dd_hh_f, mrk = marker_dd)

  # Define a window around the focal SNP for plotting
  window <- 500000  # 500 kb on either side
  xlims <- c(focal_pos - window, focal_pos + window)

  par(mfrow = c(2, 1))
  plot(furc_nn,
       xlim = xlims,
       main = paste0("Haplotype bifurcation — NN (pos ", focal_pos, ")"),
       col = c("steelblue", "coral"))
  plot(furc_dd,
       xlim = xlims,
       main = paste0("Haplotype bifurcation — DD (pos ", focal_pos, ")"),
       col = c("steelblue", "coral"))
  par(mfrow = c(1, 1))

} else {
  cat("Focal position not found in one or both filtered datasets.\n")
  cat("Trying the next-best hit instead...\n")

  # Try second-best hit
  for (i in 2:10) {
    alt_hit <- xpehh_tbl %>% arrange(desc(LOGPVALUE)) %>% slice(i)
    focal_pos_alt <- alt_hit$POSITION
    marker_nn <- which(nn_hh_f@positions == focal_pos_alt)
    marker_dd <- which(dd_hh_f@positions == focal_pos_alt)
    if (length(marker_nn) > 0 & length(marker_dd) > 0) {
      cat("Using hit #", i, "at position", focal_pos_alt, "\n")
      focal_pos <- focal_pos_alt

      furc_nn <- calc_furcation(nn_hh_f, mrk = marker_nn)
      furc_dd <- calc_furcation(dd_hh_f, mrk = marker_dd)

      window <- 500000
      xlims <- c(focal_pos - window, focal_pos + window)

      par(mfrow = c(2, 1))
      plot(furc_nn,
           xlim = xlims,
           main = paste0("Haplotype bifurcation — NN (pos ", focal_pos, ")"),
           col = c("steelblue", "coral"))
      plot(furc_dd,
           xlim = xlims,
           main = paste0("Haplotype bifurcation — DD (pos ", focal_pos, ")"),
           col = c("steelblue", "coral"))
      par(mfrow = c(1, 1))
      break
    }
  }
}
```

## Haplotype length around top hit

```{r haplen, fig.height=8}
if (exists("furc_nn") & exists("furc_dd")) {
  haplen_nn <- calc_haplen(furc_nn)
  haplen_dd <- calc_haplen(furc_dd)

  par(mfrow = c(2, 1))
  plot(haplen_nn, main = paste0("Haplotype lengths — NN (pos ", focal_pos, ")"),
       col = c(alpha("steelblue", 0.5), alpha("coral", 0.5)))
  plot(haplen_dd, main = paste0("Haplotype lengths — DD (pos ", focal_pos, ")"),
       col = c(alpha("steelblue", 0.5), alpha("coral", 0.5)))
  par(mfrow = c(1, 1))
}
```

# Sliding Window Summary

To visualise broad genomic patterns, we can compute windowed averages of |iHS|
and XP-EHH.

```{r windowed-summary, fig.height=16}
window_size <- 100000  # 100 kb windows

# Combined iHS windows
all_ihs_tbl <- all_ihs$ihs %>%
  as_tibble() %>%
  filter(!is.na(IHS)) %>%
  mutate(window = floor(POSITION / window_size) * window_size) %>%
  group_by(window) %>%
  summarise(
    mean_abs_iHS = mean(abs(IHS), na.rm = TRUE),
    max_logp = max(LOGPVALUE, na.rm = TRUE),
    n_snps = n(),
    .groups = "drop"
  )

# NN iHS windows
nn_ihs_tbl <- nn_ihs$ihs %>%
  as_tibble() %>%
  filter(!is.na(IHS)) %>%
  mutate(window = floor(POSITION / window_size) * window_size) %>%
  group_by(window) %>%
  summarise(
    mean_abs_iHS = mean(abs(IHS), na.rm = TRUE),
    max_logp = max(LOGPVALUE, na.rm = TRUE),
    n_snps = n(),
    .groups = "drop"
  )

# DD iHS windows
dd_ihs_tbl <- dd_ihs$ihs %>%
  as_tibble() %>%
  filter(!is.na(IHS)) %>%
  mutate(window = floor(POSITION / window_size) * window_size) %>%
  group_by(window) %>%
  summarise(
    mean_abs_iHS = mean(abs(IHS), na.rm = TRUE),
    max_logp = max(LOGPVALUE, na.rm = TRUE),
    n_snps = n(),
    .groups = "drop"
  )

# XP-EHH windows
xpehh_win <- xpehh_tbl %>%
  filter(!is.na(XPEHH_DD_NN)) %>%
  mutate(window = floor(POSITION / window_size) * window_size) %>%
  group_by(window) %>%
  summarise(
    mean_xpehh = mean(XPEHH_DD_NN, na.rm = TRUE),
    max_logp = max(LOGPVALUE, na.rm = TRUE),
    n_snps = n(),
    .groups = "drop"
  )

# Plot
par(mfrow = c(4, 1), mar = c(4, 4, 3, 1))

plot(all_ihs_tbl$window / 1e6, all_ihs_tbl$mean_abs_iHS,
     type = "l", col = "grey30", lwd = 1.5,
     xlab = "Position (Mb)", ylab = "Mean |iHS|",
     main = "Windowed |iHS| — All samples combined (100 kb windows)")
abline(h = 2, col = "red", lty = 2)

plot(nn_ihs_tbl$window / 1e6, nn_ihs_tbl$mean_abs_iHS,
     type = "l", col = "steelblue", lwd = 1.5,
     xlab = "Position (Mb)", ylab = "Mean |iHS|",
     main = "Windowed |iHS| — NN population (100 kb windows)")
abline(h = 2, col = "red", lty = 2)

plot(dd_ihs_tbl$window / 1e6, dd_ihs_tbl$mean_abs_iHS,
     type = "l", col = "darkorange", lwd = 1.5,
     xlab = "Position (Mb)", ylab = "Mean |iHS|",
     main = "Windowed |iHS| — DD population (100 kb windows)")
abline(h = 2, col = "red", lty = 2)

plot(xpehh_win$window / 1e6, xpehh_win$mean_xpehh,
     type = "l", col = "purple", lwd = 1.5,
     xlab = "Position (Mb)", ylab = "Mean XP-EHH",
     main = "Windowed XP-EHH — DD vs NN (100 kb windows)")
abline(h = 0, col = "grey40")
abline(h = c(-2, 2), col = "red", lty = 2)

par(mfrow = c(1, 1))
```

# Bifurcation Diagrams at iHS Peak Regions

Bifurcation diagrams show how haplotypes branch away from a focal SNP.
Under selection, one haplotype extends much further than the other, indicating
a long intact haplotype that has swept to high frequency.

We plot bifurcation diagrams for each peak in the windowed |iHS| plots:

- **Combined (ALL)**: 2 peaks (~11.4 Mb, ~12.1 Mb)
- **NN**: 5 peaks (~5.7, ~11.0, ~11.9, ~17.4, ~21.5 Mb)
- **DD**: 1 peak (~11.4 Mb)

```{r ihs-peak-bifurcation-setup}
# Helper function: find the SNP with highest |iHS| in a window, then plot bifurcation
plot_bifurcation_at_peak <- function(hh_obj, ihs_result, peak_window,
                                     pop_label, pop_col, win_size = 100000,
                                     flank = 500000) {
  # Find best SNP in window
  peak_snps <- ihs_result$ihs %>%
    as_tibble() %>%
    filter(POSITION >= peak_window, POSITION < peak_window + win_size,
           !is.na(IHS)) %>%
    arrange(desc(abs(IHS))) %>%
    slice(1)

  if (nrow(peak_snps) == 0) {
    cat(sprintf("  No SNPs found in window %.1f Mb for %s\n", peak_window / 1e6, pop_label))
    return(invisible(NULL))
  }

  focal <- peak_snps$POSITION[1]
  mrk_idx <- which(hh_obj@positions == focal)

  if (length(mrk_idx) == 0) {
    cat(sprintf("  Focal SNP %d not in haplohh object for %s\n", focal, pop_label))
    return(invisible(NULL))
  }

  furc <- calc_furcation(hh_obj, mrk = mrk_idx)
  xlims <- c(focal - flank, focal + flank)

  plot(furc, xlim = xlims,
       main = sprintf("%s — bifurcation at %s (%.3f Mb, |iHS|=%.2f)",
                      pop_label, format(focal, big.mark = ","),
                      focal / 1e6, abs(peak_snps$IHS[1])),
       col = c(pop_col, "coral"))

  return(invisible(list(furc = furc, focal = focal, ihs = peak_snps$IHS[1])))
}
```

## Combined (ALL) iHS peaks {.tabset}

### Peak 1: ~11.4 Mb

```{r bifurc-all-peak1, fig.height=5}
plot_bifurcation_at_peak(all_hh_f, all_ihs, 11400000, "ALL", "grey30")
```

### Peak 2: ~12.1 Mb

```{r bifurc-all-peak2, fig.height=5}
plot_bifurcation_at_peak(all_hh_f, all_ihs, 12100000, "ALL", "grey30")
```

## NN iHS peaks {.tabset}

### Peak 1: ~5.7 Mb

```{r bifurc-nn-peak1, fig.height=5}
plot_bifurcation_at_peak(nn_hh_f, nn_ihs, 5700000, "NN", "steelblue")
```

### Peak 2: ~11.0 Mb

```{r bifurc-nn-peak2, fig.height=5}
plot_bifurcation_at_peak(nn_hh_f, nn_ihs, 11000000, "NN", "steelblue")
```

### Peak 3: ~11.9 Mb

```{r bifurc-nn-peak3, fig.height=5}
plot_bifurcation_at_peak(nn_hh_f, nn_ihs, 11900000, "NN", "steelblue")
```

### Peak 4: ~17.4 Mb

```{r bifurc-nn-peak4, fig.height=5}
plot_bifurcation_at_peak(nn_hh_f, nn_ihs, 17400000, "NN", "steelblue")
```

### Peak 5: ~21.5 Mb

```{r bifurc-nn-peak5, fig.height=5}
plot_bifurcation_at_peak(nn_hh_f, nn_ihs, 21500000, "NN", "steelblue")
```

## DD iHS peak

```{r bifurc-dd-peak1, fig.height=5}
plot_bifurcation_at_peak(dd_hh_f, dd_ihs, 11400000, "DD", "darkorange")
```

# Candidate Region Summary

## Identify outlier regions

```{r candidate-regions}
# Define outlier thresholds
ihs_logp_thresh <- 4  # -log10(p) > 4
xpehh_logp_thresh <- 4

# Combined iHS outlier regions
all_outliers <- all_ihs$ihs %>%
  as_tibble() %>%
  filter(LOGPVALUE > ihs_logp_thresh) %>%
  arrange(POSITION) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3))

cat("=== Combined iHS outliers (-log10(p) >", ihs_logp_thresh, ") ===\n")
cat("Number of outlier SNPs:", nrow(all_outliers), "\n\n")
if (nrow(all_outliers) > 0) {
  all_outliers <- all_outliers %>%
    mutate(gap = c(0, diff(POSITION)) > 200000,
           region = cumsum(gap) + 1)
  all_regions <- all_outliers %>%
    group_by(region) %>%
    summarise(
      start_Mb = round(min(POSITION) / 1e6, 3),
      end_Mb   = round(max(POSITION) / 1e6, 3),
      n_snps   = n(),
      max_logp = round(max(LOGPVALUE), 2),
      .groups  = "drop"
    )
  print(all_regions)
}

# NN iHS outlier regions
nn_outliers <- nn_ihs$ihs %>%
  as_tibble() %>%
  filter(LOGPVALUE > ihs_logp_thresh) %>%
  arrange(POSITION) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3))

cat("\n=== NN iHS outliers (-log10(p) >", ihs_logp_thresh, ") ===\n")
cat("Number of outlier SNPs:", nrow(nn_outliers), "\n\n")
if (nrow(nn_outliers) > 0) {
  # Cluster into regions (SNPs within 200kb of each other)
  nn_outliers <- nn_outliers %>%
    mutate(gap = c(0, diff(POSITION)) > 200000,
           region = cumsum(gap) + 1)
  nn_regions <- nn_outliers %>%
    group_by(region) %>%
    summarise(
      start_Mb = round(min(POSITION) / 1e6, 3),
      end_Mb   = round(max(POSITION) / 1e6, 3),
      n_snps   = n(),
      max_logp = round(max(LOGPVALUE), 2),
      .groups  = "drop"
    )
  print(nn_regions)
}

# DD iHS outlier regions
dd_outliers <- dd_ihs$ihs %>%
  as_tibble() %>%
  filter(LOGPVALUE > ihs_logp_thresh) %>%
  arrange(POSITION) %>%
  mutate(POSITION_Mb = round(POSITION / 1e6, 3))

cat("\n=== DD iHS outliers (-log10(p) >", ihs_logp_thresh, ") ===\n")
cat("Number of outlier SNPs:", nrow(dd_outliers), "\n\n")
if (nrow(dd_outliers) > 0) {
  dd_outliers <- dd_outliers %>%
    mutate(gap = c(0, diff(POSITION)) > 200000,
           region = cumsum(gap) + 1)
  dd_regions <- dd_outliers %>%
    group_by(region) %>%
    summarise(
      start_Mb = round(min(POSITION) / 1e6, 3),
      end_Mb   = round(max(POSITION) / 1e6, 3),
      n_snps   = n(),
      max_logp = round(max(LOGPVALUE), 2),
      .groups  = "drop"
    )
  print(dd_regions)
}

# XP-EHH outlier regions
xpehh_outliers <- xpehh_tbl %>%
  filter(LOGPVALUE > xpehh_logp_thresh) %>%
  arrange(POSITION)

cat("\n=== XP-EHH outliers (-log10(p) >", xpehh_logp_thresh, ") ===\n")
cat("Number of outlier SNPs:", nrow(xpehh_outliers), "\n\n")
if (nrow(xpehh_outliers) > 0) {
  xpehh_outliers <- xpehh_outliers %>%
    mutate(gap = c(0, diff(POSITION)) > 200000,
           region = cumsum(gap) + 1)
  xpehh_regions <- xpehh_outliers %>%
    group_by(region) %>%
    summarise(
      start_Mb  = round(min(POSITION) / 1e6, 3),
      end_Mb    = round(max(POSITION) / 1e6, 3),
      n_snps    = n(),
      max_logp  = round(max(LOGPVALUE), 2),
      mean_xpehh = round(mean(XPEHH_DD_NN), 2),
      direction = ifelse(mean(XPEHH_DD_NN) > 0, "Selection in DD", "Selection in NN"),
      .groups   = "drop"
    )
  print(xpehh_regions)
}
```

# Export Results

```{r export}
# iHS results — combined and per-population
write_tsv(all_ihs$ihs %>% as_tibble(), file.path(out_dir, "chr3_ALL_iHS.tsv"))
write_tsv(nn_ihs$ihs %>% as_tibble(), file.path(out_dir, "chr3_NN_iHS.tsv"))
write_tsv(dd_ihs$ihs %>% as_tibble(), file.path(out_dir, "chr3_DD_iHS.tsv"))

# XP-EHH results
write_tsv(xpehh_tbl, file.path(out_dir, "chr3_XPEHH_DD_vs_NN.tsv"))

# Rsb results
write_tsv(rsb %>% as_tibble(), file.path(out_dir, "chr3_Rsb_DD_vs_NN.tsv"))

cat("Results exported to:", out_dir, "\n")
cat("Files:\n")
cat("  - chr3_ALL_iHS.tsv (combined NN + DD)\n")
cat("  - chr3_NN_iHS.tsv\n")
cat("  - chr3_DD_iHS.tsv\n")
cat("  - chr3_XPEHH_DD_vs_NN.tsv\n")
cat("  - chr3_Rsb_DD_vs_NN.tsv\n")
```

# Session Info

```{r session-info}
sessionInfo()
```
